---
title: "The importance of data analysis"
author: "Mark Sterken"
date: "2026-02-04"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
editor: visual
execute:
  warning: false
  message: false
---

## The importance of data analysis

### Introduction to day 1

Today we will focus on goal 1 (Built basic knowledge on using R and R studio) and goal 5 \[Apply the data-science cycle (load, inspect, ~~clean~~, analyse, present)\].

In order to illustrate some concepts, we will make use of data from a paper on nematode infections in *Arabidopsis thaliana*[@Warmerdam2018].

::: {.callout-tip appearance="simple"}
### *M. incognita* infections in *A. thaliana*

Plant susceptibility to nematodes is a complex trait, with a lot of variation in susceptibility between individuals of the same species.
At [the Laboratory of Nematology](https://www.wur.nl/en/chair-groups/laboratory-of-nematology), we study this plant-property.
The study by Sonja Warmerdam (who, before her PhD, studied plant science)[@Warmerdam2018] was conducted on 340 ecotypes of the model plant *A. thaliana*.
These were infected with the tropical root knot nematode *Meloidogyne incognita*.
This plant-parasitic nematode is the cause of a major agricultural burden globally[@Jones2013].
We will use data from the initial screening (that took over 1.5 years) that was conducted for this research project.

![Fig. 1 Quantitative variation in susceptibility of *Arabidopsis thaliana* to the root-knot nematode *Meloidogyne incognita*. Boxplot of the number of egg masses per plant. The thick bar indicates the median and the box indicates the 25 and 50 percent quantiles. The dots represent the individual observations on 340 ecotypes of Arabidopsis (accessions) at 6 wk after inoculation with 2^nd^ stage juveniles of *M. incognita*. The black boxplot and dots indicate the number of egg masses per plant for Col-0, which was used as a reference throughout this study.](Week1_day1_Warmerdam.png)
:::

As you gathered by now, this is a course in which we will use the versatile, statistical programming language `R`.
We will teach you how to use R via RStudio.
In particular we will teach you how to write and document code for data analysis.
By the end of this week, you will have made your first steps towards making documents as you are reading now: [Quarto markdown](https://quarto.org/docs/guide/).


### Set up your R environment

In this section of the book, the basics of setting up RStudio are explained. You are welcome to revisit this section any day of the course.


::: {.callout-note appearance="simple"}
### A short intro to RStudio and scripts

When you open RStudio, you see four panes.

![The Rstudio layout with four panes: (1) On the top-left a text editor to write and document your code in. (2) On the top-right a pane that - for instance - can show which objects are available in the environment. (3) On the bottom-left the console which can execute your code. (4) On the bottom-right a pane where, among others, you can: browse through files, plots, and get an overview of packages.](Week1_day1_RStudio_layout.png)
:::

Now, check the message you see in the console.The R version displayed is important.

::: {#exr-Rversion}
### Getting to grips with R
Start RStudio and browse through to find out:
1. Which version of R are you running?
2. Where can you find which packages are activated?
3. Which version of the stats package is active?
4. Where on your Harddrive is the current (default) working directory?

:::

The scripts you are using the first week should work with R versions 4.3 and onwards.
If your version number is lower, please first shut-down RStudio and install a newer version of R.
You can find out how to do so [here](https://support.wur.nl/esc?id=kb_article&sysparm_article=KB0010713&table=kb_knowledge&searchTerm=r%20installation). RStudio should automatically pick up on the newer installed version. You can check if this worked by checking the version number of R after installation.

It is generally advisable to install a new version of R at the start of a new project (e.g. BSc thesis).

#### Documenting your code

Your analysis starts with making a file to document your progress and thereafter you follow the standard steps in data analysis

1.  Loading
2.  Inspecting
3.  Cleaning
4.  Analyzing
5.  Presenting

For the tutorials, assignments, and exams, it's important that you built a file documenting your code. On the first two days, we will provide a full-scale template for you, that you below or - alternatively - you can download on [Brightspace](https://brightspace.wur.nl/). which you can copy paste into the text editor in R-studio.

```{r w1d1_script, eval=FALSE, class.source = 'fold-show'}

# BIF20806 - Day 1 ---------------------------------



# set a work directory ----------------------------------

###this function you can use to automatically set a location
#setwd("D:/")


# Install packages ----------------------------------------------------


# Load data -----------------------------------------------------------


# Inspect the data ----------------------------------------------------


# Clean data ----------------------------------------------------------
   

# Analyzing -----------------------------------------------------------
        
    ### Normal distribution ###########################################


    ### T-test #########################################################


# Presenting ----------------------------------------------------------

    ### Simple box-plot ###############################################



# For-loop ----------------------------------------------


```

::: {#exr-SetTemplate}
#### Prepare your documentation
In RStudio go to File > New File > Quarto document...

Paste the template in and fill in your own information. 

This type of document is also what you will make for the assignments. Furthermore, you will also make your exam based on this template.

:::

#### Telling R where on the hard drive it should work from.

First, you need to set a work directory (tell R where on the hard drive you'll be working from).
This can be done with the function `setwd()`, where you have to fill in the work directory yourself.
If you want to know how to use that function type `?setwd()` into the console.

::: {#exr-WD}
#### Set your work directory

Run the script below to get information on the `setwd()` function. thereafeter, use it to set the location where R should work from. It is advisable to make a folder on your HDD for items related to this course.

```{r w1d1_set_working_directory, eval=FALSE, class.source = 'fold-show'}
    # Set a work directory
    ?setwd()

    ###Now type the function to set a location on your HDD to work from

```
:::

Alternatively, you can also use the three dots `...` in the top-right corner of the files pane of RStudio to navigate towards the location you will be working from.

#### Installing packages.

Typically R-users use several 'packages'. These packages, give extra functionality to R. The main repositories are:

1.  [CRAN](https://cran.r-project.org/)
2.  [Bioconductor](http://www.bioconductor.org/)

In this tutorial we will be making use of various packages. During week 1 we will make use of the {readxl} and {tidyverse} package each day, including during the exam. By the end of this first week, we expect you are able to install and activate packages.

::: {#exr-install_packages}
#### Install packages
Install the {readxl} and {tidyverse} packages.

```{r install_packages_readxl_tidyverse, eval=FALSE, class.source = 'fold-show'}
    # Install the package
    install.packages("readxl")    # For reading Excel files
    install.packages("tidyverse") # For wrangling data and generating plots
```
:::

Note that you only need to install a package once, and only re-install it if the package has a new version or you install a new version of R.

::: {#exr-activate_packages}
#### Activate packages
To activate the package, we need to load it into memory:

```{r w1d1_activate_packages_readxl_tidyverse, eval=FALSE, class.source = 'fold-show'}
    # Activate the package
    library("readxl")    
    library("tidyverse") 
```
:::

Now you are all set to do data-analysis in R!

## Analysis of quantitative phenotypic data from an experiment

### Loading data into R

During this course, we will load different types of data files into R.
This is an essential skill for working with real-world data.
In the following days we'll cover a couple of file-types that are common.
For today, we have provided you with data in tab-delimited text format.
Try to load this data using the funcion `read.delim()`.

```{r w1d1_load_tabdelim_ASSIGNMENT, include = TRUE, eval=FALSE}

    # Replace with the path to your csv file
    pheno_data <- read.delim("./data/BIF20806_Warmerdam_dataset.txt")

```

### Inspecting data

There are a couple of handy functions you should use to check the data you loaded: `head()`, `tail()`, and `summarise()`.

```{r w1d1_check_the_data, include = TRUE, eval=FALSE, class.source = 'fold-show'}

    # Check the first 6 rows
    head(pheno_data)

    # Check the last 6 rows
    tail(pheno_data)
    
    # get a summary of the data
    summary(pheno_data)
    
```

As you see above, these functions give a good overview of the dataset.
It is important to do these basic checks in order to confirm that your data is in the right shape.

You can also specifically check items in an object by using `[]`

```{r w1d1_check_the_data_2, include = TRUE,eval=FALSE, class.source = 'fold-show'}

    # The 18th row
    pheno_data[18,]

    # The 3rd column
    pheno_data[,3]
    
    # Or combined the 18th row of the 3rd column
    pheno_data[18,3]
    
    # the 18th value of the Line column; as a vector is one-dimensional only 1 coordinate is needed.
    pheno_data$Line[18]
    
```

### Inspect and clean data using ggplot2 and tidyverse

To actually clean data, you need to inspect it (take a peak).
I typically make a couple of plots.
For example, here using ggplot (comes with the tidyverse package).
You can read more about the plotting functions on their [website](https://ggplot2.tidyverse.org/)

#### Boxplots and histograms

The code below produces a boxplot, great for checking continuous data.
I notice in the plot that there are a couple of single points (the horizontal stripes) per strain, also, there seem to be a couple of really high values.
Let's check these.

```{r w1d1_make_a_ggplot2_boxplot, include = TRUE,eval=FALSE, class.source = 'fold-show'}

        ggplot2::ggplot(pheno_data,aes(x=genotype,y=eggmass)) +
        geom_boxplot()
    
```

Let's have a look at the distribution of the data.
Try to plot a histogram, for that you need to adjust the code above a bit.
You need to use the `geom_histogram()` function.
Use the [ggplot2 website](https://ggplot2.tidyverse.org/) to figure out how to make this plot.

```{r w1d1_make_a_ggplot2_histogram_ASSIGNMENT, include = TRUE,eval=FALSE}

        ggplot2::ggplot(pheno_data,aes(x=eggmass)) +
        geom_histogram()
    
```

#### Use table to inspect observation per genotype

For the single values per genotype, it can be handy to use the `table()` function.
Try to use this function on the column *genotype*.
You can use `pheno_data$genotype` to access that column.

```{r w1d1_table_for_genotypes_ASSIGNMENT, include = TRUE,eval=FALSE}

    ?table()

    table(pheno_data$genotype)

```

This way, you can check the number of observations per genotype.

Now use `table()` to check the Line names.

```{r w1d1_table_for_lines_ASSIGNMENT, include = TRUE,eval=FALSE}

    table(pheno_data$Line)

```

#### Cleaning and preparing the data

In this example, all data is clean and we will not spend time cleaning it up.
But it is important to realize that there typically are errors and inconsistencies in datasets.
Figuring these out and fixing them is a large chunk of doing data analysis.

### Analyzing

#### Normal distribution

Let's figure out whether the data is normally distributed.
For this we can use the functions `stat_qq()` and `stat_qq_line()`.

```{r w1d1_qqplot_ASSIGNMENT, include = TRUE,eval=FALSE}

    ggplot(pheno_data,aes(sample=eggmass)) + 
    stat_qq() + stat_qq_line()

```

Alternatively, you can use a Shapiro-Wilk test (`shapiro.test`), this tests the deviation from normality.

```{r w1d1_shapiro_ASSIGNMENT, include = TRUE,eval=FALSE}

    shapiro.test(pheno_data$eggmass)

```

what does the p-value mean?

#### T-test

Now we can use a t-test to compare between two samples.
For that, we first need to generate a smaller set of genotypes.
We want to test a genotype versus the *A. thaliana* reference genotype Col-0.
Let's take the also popular Cvi-0 genotype to compare.

In the lines below, we first group (`dplyr::group()`) the batches together (screenings were done at a particular date), then we keep only the screenings with Col-0 and Cvi-0 in them (`dplyr::filter()`).
Thereafter we remove the grouping (`dplyr::ungroup()`) and keep only the Col-0 and Cvi-0 observations (again with `dplyr::filter()`).

```{r w1d1_filter_Warmerdam, include = TRUE,eval=FALSE, class.source = 'fold-show'}

    data.test <- dplyr::group_by(pheno_data,screening) %>%
                 dplyr::filter(("Col-0" %in% Line & "Cvi-0" %in% Line)) %>%
                 dplyr::ungroup() %>%
                 dplyr::filter(Line %in% c("Col-0","Cvi-0"))

```

Now we can test.
See if you can test whether these two *A. thaliana* genotypes are different in their susceptiblity to *Meloidogyne incognita*.
Use `t.test()`

```{r w1d1_ttest_Warmerdam_ASSIGNMENT, include = TRUE,eval=FALSE}
t.test(eggmass~Line,data=data.test)
###you should have p-value = 2.588e-06
```

### Presenting

Can you now also make a plot of what you just tested?
Use `geom_boxplot()` to visualize the distribution and `geom_jitter()` to show the individual measurements.

```{r w1d1_ttestplot_Warmerdam_ASSIGNMENT, include = TRUE,eval=FALSE}

ggplot(data.test,aes(x=Line,y=eggmass)) +
geom_boxplot() + geom_jitter(height=0,width=0.25)
```

We can make it a bit fancier by plotting the p-value as well.
Try to include it into the plot by using the function `annotate()`.

```{r w1d1_ttestplot_pval_Warmerdam_ASSIGNMENT, include = TRUE,eval=FALSE}

###I'm getting the statistics here
statplot <- broom::tidy(t.test(eggmass~Line,data=data.test))

p1 <- ggplot(data.test,aes(x=Line, y=eggmass)) +
      geom_boxplot() + 
      geom_jitter(height=0, width=0.25) +
      annotate("text", x=1.5, y=max(data.test$eggmass), label=paste("p =", signif(statplot$p.value, 2)))
```

#### Saving plots

Now you have your plot, you need to save it.
This also requires attention.
There are multiple 'graphical devices' in `R` that you can use to plot.

The function I use most often in `pdf()`.
The reason is that this function produces vectorized pdf's, which means that figures have infinite resolution.
When you want to save your plot using this function, either save the figure in a file (as in the previous section), or plot directly within the 'graphical device', if you use {ggplot2}, you often have to explicitly define that you print the plot, by adding the function `print()`.
Your entire plotting code than should be included in the brackets.

I also need to define how wide and high I want to make the plot (in inches).
I play around with the dimensions a lot, to get the figure in the right proportions.
Note that the plotting is ended with `dev.off()`.
This shuts down the graphical device.
If you do not do that, it remains open and you will not see any plots appear on your plotting window in Rstudio.

```{r w1d1_pdf_of_warmerdam_plot, include = TRUE, eval=FALSE, class.source = 'fold-show'}

pdf(file="Figure_Col0-Cvi0.pdf", width=3, height=4)
print(p1)
dev.off()   
        
```

### A for-loop

#### The basics of a loop

Within the data there are 332 unique genotypes.
Verify this using the function `unique()` to get the unique genotypes and `length()` to count.

```{r w1d1_check_length_ASSIGNMENT, include = TRUE,eval=FALSE}

genotypes <- unique(pheno_data$Line)

length(genotypes)

```

What if we want to plot all of these versus Col-0?
Writing all these plots line-by-line is not doable.
For this we can use a for-loop.
with the function `for()` you can start a loop.
The `for()` function counts along a variable.
There are also other loops possible, for instance with `while()` that continues if a condition is met.
Let's first look at a simple for-loop.

```{r w1d1_for_loop_check, include = TRUE, eval=FALSE, class.source = 'fold-show'}

for(i in 1:30){
    cat("",i)
}

```

That probably went too fast, but it printed the values 1:30 in your console.
You can loop over any vector:

```{r w1d1_for_loop_check2, include = TRUE, eval=FALSE, class.source = 'fold-show'}

for(i in c("N. americanus","E. vermicularis","M. incognita","A. suum",
           "T. semipenetrans","O. volvulvus","D. dipsaci","E. arcuatus")){
    print(i)
}

```

#### For-loop over t-tests

Now, let's automatize the t-test function.
This is a bit simpler than plotting.
For this, you need to make a for-loop cycling over the genoypes in `pheno_data`.
You can remove Col-0 from the vector using `genotypes <- genotypes[genotypes != "Col-0"]`.
Now, with what you read above, write a for-loop doing the t-test of each genotype versus Col-0 and make sure to pint the values in the console (either use `cat()` or `print()`).

```{r w1d1_for_loop_t-test_ASSIGNMENT, include = TRUE,eval=FALSE}

genotypes <- unique(pheno_data$Line)
genotypes <- genotypes[genotypes != "Col-0"]

output <- as.list(NULL)
for(i in 1:length(genotypes)){
  data.test <- dplyr::group_by(pheno_data,screening) %>%
               dplyr::filter(("Col-0" %in% Line & genotypes[i] %in% Line)) %>%
               dplyr::ungroup() %>%
               dplyr::filter(Line %in% c("Col-0",genotypes[i]))
  
  cat("",t.test(eggmass~Line,data=data.test)$p.value)
}

```

#### For loop for plotting

Now we can add a plot, for that we need to save the plots in an object (like we did with p1 before).
Now, we will make a specific type of object: a `list`.
Note that this `list` is not what you think of as a list.

`Lists` are objects that can contain various other types of objects.
For instance, I can place a `vector` or a `dataframe` into a `list`.
You can access a location in a list using `[[]]`.

```{r w1d1_a_list, include = TRUE,eval=FALSE, class.source = 'fold-show'}

###this makes an empty list
output <- as.list(NULL)

###you can add the first item
output[[1]] <- c("N. americanus","E. vermicularis","M. incognita","A. suum",
                 "T. semipenetrans","O. volvulvus","D. dipsaci","E. arcuatus")

###and a second item
output[[2]] <- matrix(runif(100),ncol=4)

```

As we saw before, you can also access specific locations within these items, simply by adding a `[]` on top of the `[[]]`.

```{r w1d1_access_a_list, include = TRUE,eval=FALSE, class.source = 'fold-show'}

# The 18th row
output[[2]][18,]

# The 3rd column
output[[2]][,3]

# Or combined the 18th row of the 3rd column
output[[2]][18,3]

# the 8th value of the Line column; as a vector is one-dimensional only 1 coordinate is needed.
output[[1]][8]
    
```

Now take the plotting function we wrote before and use it to save plots into an object.
Note that this takes a while to run (\~2 minutes).

```{r w1d1_for-loop_Warmerdam, include = TRUE,eval=FALSE}

genotypes <- unique(pheno_data$Line)
genotypes <- genotypes[genotypes != "Col-0"]

output <- as.list(NULL)
for(i in 1:length(genotypes)){

data.test <- dplyr::group_by(pheno_data,screening) %>%
             dplyr::filter(("Col-0" %in% Line & genotypes[i] %in% Line)) %>%
             dplyr::ungroup() %>%
             dplyr::filter(Line %in% c("Col-0",genotypes[i])) %>%
             dplyr::mutate(Line = factor(Line,levels=c("Col-0",genotypes[i])))

statplot <- broom::tidy(t.test(eggmass~Line,data=data.test))

output[[i]] <- ggplot(data.test,aes(x=Line,y=eggmass)) +
               geom_boxplot() + geom_jitter(height=0,width=0.25) +
               annotate("text",x=1.5,y=max(data.test$eggmass),label=paste("p =",signif(statplot$p.value,2))) +
               theme_bw() + 
               labs(x=expression(bolditalic("A. thaliana")~bold("genotype")),y=expression(bolditalic("M. incognita")~bold("eggmasses (n)")))
}
```

#### Save the plots

We can save the plots in various ways.
We can write one big pdf that contains all the plots.

```{r w1d1_pdf_of_warmerdam_plot_all_vs_Col-0, include = TRUE, eval=FALSE, class.source = 'fold-show'}

pdf(file="Figure_All_Col0-comparisons.pdf",width=3,height=4)
    for(i in 1:length(output)){
        print(output[[i]])
    }
dev.off()   
        
```

or we can make a png of each plot individual.

```{r w1d1_png_of_warmerdam_plot_all_vs_Col-0, include = TRUE, eval=FALSE, class.source = 'fold-show'}

for(i in 1:length(output)){
    png(file=paste("Figure_Col0-",genotypes[i],".png",sep=""),width=3,height=4,units="in",res=300)
        print(output[[i]])
    dev.off()
}
        
```
