---
title: "Summary statistics and functions"
format: html
---

## Introduction

A simple but effective way of starting to explore a dataset is to compute a few summary statistics. In the previous week we already encountered a few summary statistics: we looked at e.g. the numbers of rows and columns of a table. In this section we will explore a few numerical summary techniques, such as computing the mean and standard deviation of a numeric variable, or counts/frequencies of a categorical variable. At the same time, we will look into one of the more powerful features of using a programming language: the possibility to create functions.

```{r}
remotes::install_github('wur-bioinformatics/planteda')
library(planteda)
data('corre_continuous')
```

## Summary statistics

Perhaps one of the simplest and most intuitive summary statistics is the mean (typically indicated with the greek letter $\mu$). The mean can be calculated as the sum of all values in a variable divided by the number of values in that variable (@eq-mean).

$$
Mean(X) = \mu = \frac{1}{N}\sum_{i=1}^{n}{x_i}
$$ {#eq-mean}

A common interpretation of the mean is that it represents the expected value of a variable. In other words: if you have no other information, the mean would be your best guess on average. In the example below this would mean that your best guess of the leaf nitrogen content of a plant, without knowing anything else about that plant, would be 22.9.

Being a statistical programming language, R comes with a default function to calculate the mean:

```{r}
# Calculate the average leaf nitrogen content
mean(corre_continuous[['leaf_N']])
```

::: {.callout-note appearance="simple"}
### Implementing the mean calculation in a custom function

Since R already comes with a default function to calculate the mean this example might seem redundant. At the same time, this gives us the opportunity to check whether our custom implementation returns the correct result.

```{r}
# Here we define a function to perform the calculation of the mean, 
# closely following the mathematical description
custom_mean <- function(input){
  total = sum(input)
  number = length(input)
  return(total / number)
}

# This should give the same result as using the default `mean` from R
custom_mean(corre_continuous[['leaf_N']])
```
:::

$$
Variance(X) = \sigma^2 = \frac{1}{N}\sum_{i=1}^{n}{(\mu-x_i)^2}
$$ {#eq-variance}


$$
StandardDeviation(X) = \sigma = \sqrt{\frac{1}{N}\sum_{i=1}^{n}{(\mu-x_i)^2}}
$$ {#eq-standard-deviation}

::: {#exr-variance}
### Implement the variance calculation in a custom function

```{r}
#| eval = FALSE
custom_variance <- function(input){
  # ... some initial calculation goes here
  variance <- # ... final calculation goes here
  return(variance)
}
```
:::
