---
author: "Mark Sterken"
date: "2026-02-11"
editor: source
execute:
  eval: false
  warning: false
  message: false
---

# QMD notebook cheatsheet {.unnumbered}

## Setting up QMD
You start a QMD with a yaml header, this is information in-between three dashes (\-\-\-). The first three dashes indicate the start of the header, the last three the end. There are some guidelines to writing yaml. But for this course it suffices to copy the header from the first two days

```{r}
#| code-fold: false

---
title: "BIF20806_week1day1_notebook"
editor: source
date: ADD
author: ADD
registration_number: ADD
format: 
  html
execute: 
  eval: false
---
  

```

## Making a code chunk in QMD
Code is written in QMD between three back ticks \`\`\`. After these backticks you should state \{r\}, which indicates that the code chunk is in the r-language. The yaml header you use states that this code is not evaluated (eval: false). You can force evaluation by either setting this to true (eval: true) or setting it to true in a specific code block using "\#\| eval: true".


## R setup in QMD
The first code chunk in the document should be \{r setup\}, this chunk specifies some general conditions. In particular, on some systems it is required to set the root directory (your workdirectory) here as well (in addition to `setwd()`).

```{r setup}
#| code-fold: false

    options(repos = c(CRAN = "https://cloud.r-project.org"))
    ### Some of you will have problems with setting the workdirectory (setwd())
    ### Here you need to set your work directory as well
    knitr::opts_knit$set(root.dir = "YOUR_WD") 
```


## Writing text in QMD

| Feature           | Syntax                        | Explanation                               |
|------------------|--------------------------|----------------------------|
| Bold              | \*\*text\*\*                  | Makes text bold                           |
| Italic            | \*text\*                      | Makes text italic                         |
| Bold + Italic     | \*\*\*text\*\*\*              | Bold and italic                           |
| Inline code       | \`code\`                      | Formats inline code                       |
| Strikethrough     | \~\~text\~\~                  | Strikes through text                      |
| Headings          | \#, \#\#, \#\#\#, \#\#\#\#    | Creates headings (H1--H4)                 |
| Bullet list       | \- item                       | Creates unordered lists                   |
| Numbered list     | 1\. item                      | Creates ordered lists                     |
| Task list         | \- \[ \] item                 | Creates checkboxes                        |
| Link              | https://example.com           | Creates a clickable link                  |
| Link              | \[text\](https://example.com) | Creates a clickable link with custom text |
| Image             | image.png                     | Inserts an image                          |
| Image             | \!\[text\](image.png)           | Inserts an image with a caption           |
| Blockquote        | \> text                       | Creates an indented quote                 |
| Horizontal rule   | \-\--                         | Inserts a horizontal line                 |
| Footnote          | text\[\^1\] / \[\^1\]: note   | Adds a footnote                           |
| Line break        | text (two spaces)             | Forces a new line                         |
| Escape characters | \*, #, \_                     | Shows literal symbols                     |




